---
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
---
## Key for workshop 8 exercises
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, cache = T)
```


## Exercises
For these exercises use the [LDWF Pontchartrain seine sampling dataset](https://github.com/BSC6926-B52_Fall2025/workshopScripts/blob/main/data/Pontchartrain.csv). This dataset is the abundance for each species (in wide format) for 16 sites over 1 year.

1. Using the Pontchartrain dataset calculate the mean and SD for species richness, Shannon, Simpson, and inverse Simpson for each sampling site.

```{r}
library(vegan)
library(tidyverse)

df = read_csv('data/Pontchartrain.csv') |> 
  mutate(richness = specnumber(across(`Skipjack Herring`:`Feather Blenny`)),
         H = diversity(across(`Skipjack Herring`:`Feather Blenny`)),
         simp = diversity(across(`Skipjack Herring`:`Feather Blenny`), "simpson"),
         invsimp = diversity(across(`Skipjack Herring`:`Feather Blenny`),"inv"))

d = df |> 
  group_by(site) |> 
  summarize(across(richness:invsimp, list(mean = mean, sd = sd)))

d
```


2. Plot the dominance (Whittiker) and K-dominance curves for each site.

```{r}
library(viridis)
marsh_w = read_csv('data/Pontchartrain.csv') |> 
  mutate(site = as.character(site),
         month = month(date),
         richness = specnumber(across(`Skipjack Herring`:`Feather Blenny`))) 

marsh_l = marsh_w |> 
  pivot_longer(cols = 3:73, 
               names_to = "Species", 
               values_to = "Count") 

df = marsh_l |> 
    group_by(site) |> 
    filter(Count > 0) |> 
    mutate(Total = sum(Count)) |> 
    group_by(site, Species) |>
    summarise(Count_Spp = sum(Count),
              Total_Count = max(Total)) |> 
    mutate(p_i = Count_Spp/Total_Count, 
           rank = length(unique(Species))-rank(p_i)) |> 
    ungroup()

ggplot(df, aes(rank, p_i, color = site))+
  geom_line(size = 1)+
  labs(x = 'Species rank',
       y = 'Dominance',
       color = 'site')+
  scale_color_viridis_d()+
  theme_bw()

## K-dominance curves
# Cumulative dominance by species rank


df = marsh_l |> 
    group_by(site) |> 
    filter(Count > 0) |> 
    mutate(Total = sum(Count)) |> 
    ungroup() |> 
    group_by(site, Species) |>
    summarise(Count_Spp = sum(Count),
              Total_Count = max(Total)) |> 
    mutate(p_i = Count_Spp/Total_Count, 
           rank = length(unique(Species))-rank(p_i)) |> 
    arrange(rank, .by_group = T) |> 
  mutate(cumsum = cumsum(p_i))

ggplot(df, aes(rank, cumsum, color = site))+
  geom_line(size = 1)+
  labs(x = 'Species rank',
       y = 'Cumulative Dominance',
       color = 'site')+
  scale_color_viridis_d()+
  theme_bw()


```


3. Plot the $\alpha$, $\beta$, and $\gamma$ diversity for each site for each sampling date in 2007. Use either additive or multiplicative $\beta$ diversity. Plot diversity over time.

```{r}
marsh_w = read_csv('data/Pontchartrain.csv') |> 
  mutate(site = as.character(site),
         month = month(date),
         richness = specnumber(across(`Skipjack Herring`:`Feather Blenny`))) 

marsh_l = marsh_w |> 
  pivot_longer(cols = 3:73, 
               names_to = "Species", 
               values_to = "Count") 

marsh_l

gammaDiv = length(unique(marsh_l$Species))

# calculate beta diversity
betaDiv = marsh_w |> 
  group_by(month) |> 
  summarise(alpha = mean(richness, na.rm = TRUE),
            gamma = gammaDiv,
            beta_a = gamma - alpha,
            beta_m = gamma/alpha) |> 
  pivot_longer(alpha:beta_m, names_to = 'metric', values_to = 'value')

# plot 

ggplot(betaDiv, aes(month, value, color= metric))+
  geom_point(size = 2)+
  geom_line(linewidth = 1)+
  scale_y_continuous(limits = c(0,75))+
  scale_color_viridis_d(labels = c(expression(alpha),
                                   expression(beta ~ 'additive'),
                                   expression(beta ~ 'multiplicative'),
                                   expression(gamma)))+
  labs(x = 'Month', y = 'Diversity', color = 'Diversity')+
  theme_bw()


```


4. Plot the Bray-Curtis and Jaccard dissimilarity for each site.

```{r}
df = read_csv('data/Pontchartrain.csv')|>
  group_by(site) |> 
  summarise(across(`Skipjack Herring`:`Feather Blenny`, mean))
  
  
marsh_comm = df |> 
  select(-site)

site = df$site

marsh.nmds.bc = metaMDS(marsh_comm, distance = "bray", k = 2, try = 100)
marsh.nmds.jc = metaMDS(marsh_comm, distance = "jaccard", k = 2, try = 100)

bc = data.frame(marsh.nmds.bc[["points"]]) |> 
  mutate(site = site, 
         Dissimilarity = 'Bray-curtis')

jc = data.frame(marsh.nmds.jc[["points"]]) |> 
  mutate(site = site, 
         Dissimilarity = 'Jaccard')

nmds_output = bind_rows(bc,jc) |> as_tibble()

ggplot(nmds_output, aes(MDS1, MDS2, color = as.factor(site)))+
  geom_point(size = 4)+ 
  facet_wrap(~Dissimilarity)+
  labs(color = 'site')+
  scale_color_viridis_d(option = 'turbo')+
  theme_bw()

```

